Unsupervised embedding of trajectories captures the latent structure of mobility

### 2020-10-01 paper/current/main.tex
performance results

We show that the embedding distance, compared with the geographical distances, predicts actual mobility flows. 
For the case of scientific mobility, the amount of variance in the expected flux explained by the embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of the geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds when considering the subset of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
We fit the gravity model given by Eq.~\ref{eq:gravity_basic} (see Data and Methods) and computed the expected flux  $\hat{T}_{ij}$ using the geographic distance and the embedding distance as parameters.
The embedding distance produces better predictions of actual flux than does the geographic distance (Fig.~\ref{fig:gravity_performance}c-d), which also holds for the subsets of only domestic and only international mobility (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).
The embedding distance defined from the cosine distance between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion-based'' network distance measures such as the personalized-page rank scores (see \SI) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}). These results are consistent in the U.S. flight itinerary data.
As with scientific mobility, the embedding approach better predicts actual flows of airline passengers than the geographic distance. 
The embedding distance explains more of the expected flux between U.S. airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}e) than does the geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}f). 
By fitting the gravity model given by Eq.~\ref{eq:gravity_basic} using each of the embedding distance and the geographic distance as parameters, and we find that predictions with embedding distance out-performed those with the geographic distance (Fig.~\ref{fig:gravity_performance}g-h). 
Our results demonstrate that the embedding distance better captures patterns of actual mobility than does the geographic distance across datasets.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts global scientific mobility and U.S flight itineraries.}
		\textbf{a-b.} Embedding distance (cosine distance between organization vectors) better explains the expected flux of researchers between organization pairs than geographic distance.
		The red line is the line of the best fit.
		Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted, but are too small to be visible.
		Correlation is calculated on the data in the log-log scale. 
		The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{c-d.} Predictions of flux between organization pairs made using embedding distance outperform those made using geographic distance.
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$;
		a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
		We show the generalizability of this approach by repeating it on a a dataset of U.S. flight itineraries. 
		\textbf{e-f.} Embedding distance better explains the expected flux of passengers between U.S. airports than does geographic distance. 
		\textbf{g-h.} Predictions of flux between airport pairs made using embedding distance outperform those made using geographic distance. 
	}
	\label{fig:gravity_performance}
\end{figure}

>>>>

We show that the embedding distance, compared with the geographical distance, predicts actual mobility flows. 
In the case flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b). 
We fit the gravity model given by Eq.~\ref{eq:gravity_basic} (see Data and Methods) and computed the expected flux  $\hat{T}_{ij}$ using the geographic distance and the embedding distance as parameters.
The embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}c-d).
These results are consistent in the case of scientific mobility. 
the amount of variance in the expected flux explained by the embedding distance ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) is again more than double that explained by the geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}f).
This pattern holds when considering the subset of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
As with flight itineraries, we fit the gravity model on the scientific mobility data using the gravity model.
Predictions made using the embedding distance outperform those made using the geographic distance (Fig.~\ref{fig:gravity_performance}g-h), which also holds for the subsets of only domestic and only international mobility (Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:predict_emb_bymodel}).
For scientific mobility, the embedding distance defined from the cosine distance between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion-based'' network distance measures such as the personalized-page rank scores (see \SI) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Our results demonstrate that the embedding distance better captures patterns of actual mobility than does the geographic distance across datasets.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts U.S flights and global scientific mobility.}
		\textbf{a-b.} Embedding distance (cosine distance between organization vectors) better explains the expected flux of passengers between U.S. airports than does geographic distance. 
		The red line is the line of the best fit.
		Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted, but are too small to be visible.
		Correlation is calculated on the data in the log-log scale. 
		The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{c-d.} Predictions of flux between airport pairs made using embedding distance outperform those made using geographic distance. 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$;
		a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
		Results are consistent in the case of scientific mobility. 
		\textbf{e-f.} Embedding distance better explains the expected flux of researchers between organization pairs than geographic distance.
		\textbf{g-h.} Predictions of flux between organization pairs made using embedding distance outperform those made using geographic distance.
	}
	\label{fig:gravity_performance}
\end{figure}
